Iterative Monomorphisation

### 2025-02-24 iterative_monomorphisation.tex
tune

Automatic theorem provers are employed to provide automation to users of proof assistants. Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle/HOL~\cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae. By contrast, many automatic provers operate only on monomorphic logics. One approach to closing this gap is to extend provers to natively support polymorphism, as has been done for Vampire~\cite{bhayat-reger-2020}. This, however, entails a lot of work that needs to be redone for every prover.

@@

Any translation method based on instantiation is
bound to be incomplete.
By a type version of the compactness theorem, we have that for any first order polymorphic formula \(\varphi\), there exists an equisatisfiable finite set of monomorphic instances of \(\varphi\). However, this set cannot be computed~\cite[Theorem~1]{expr-poly-types}.

@@

Consider the unary type constructor \ty{list}. If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, it may be possible to generate the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc. However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has been generated. This example also makes it clear that the set of instances that can be generated is infinite.

>>>>

Automatic theorem provers are employed to provide automation to users of proof assistants. Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle\slash HOL \cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae. By contrast, many automatic provers operate only on monomorphic logics. One approach to closing this gap is to extend provers to natively support polymorphism, as has been done for Vampire~\cite{bhayat-reger-2020}. This, however, entails a lot of work that needs to be redone for every prover.

@@

Any translation method based on instantiation is
bound to be incomplete.
By a type version of the compactness theorem, there exists for any first order polymorphic formula \(\varphi\) an equisatisfiable finite set of monomorphic instances of \(\varphi\). However, this set cannot be computed~\cite[Theorem~1]{expr-poly-types}.

@@

Consider the unary type constructor \ty{list}. If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc., can be generated. However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has been generated. This example also makes it clear that the set of instances that can be generated is infinite.
